\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-03-04 misc/JSS article/jss-article-rnw/article.Rnw
Implementing JSS form comments: commas after e.g. and i.e.

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g. \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g. \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

What is the purpose of combining all of this in a package? The short answer is to make computations in \proglang{R} as flexible and powerful as possible. The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g. by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline

@@

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars} and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy} or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g. surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.}. Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function. However, as demonstrated below, the possibility of combining vectorized statistical functions also permits calculating more complex statistics in a vectorized way. \newline

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e. simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of Single Instruction Multiple Data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}. The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there, not here for parsimony) demonstrate their basic usage.

@@

fmean(mtcars$mpg)                                  # Vector
fmean(EuStockMarkets)                              # Matrix
fmean(mtcars)                                      # Data Frame
fmean(mtcars$mpg, w = mtcars$wt)                   # Weighted mean
fmean(mtcars$mpg, g = mtcars$cyl)                  # Grouped mean
fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)   # Weighted group mean
fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt) # Of data frame
fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")    # Replace data by g. mean

@@

  3 && \code{"-"}   && subtract (i.e. center) \\

@@

  5 && \code{"/"}   && divide (i.e. scale) \\
  6 && \code{"\%"}     && compute percentages (i.e. divide and multiply by 100) \\

@@

  9 && \code{"\%\%"} && modulus (i.e. remainder from division by  \code{STATS}) \\
  10 && \code{"-\%\%"} && subtract modulus (i.e. make data divisible by \code{STATS})

@@

to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can be an \class{indexed\_frame} with any number of individual and/or time identifiers (e.g. an indexed \class{data.table} is fully functional for other operations). The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}. The basic syntax is:

@@

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \code{base::table}. It is enhanced both in a statistical and computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights - vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
%
<<qtab>>=
library(magrittr) # World after 2015 (latest country data)
wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
wlda15 %$% qtab(OECD, income)
@
This shows the total population (latest post-2015 estimates) in millions.
<<qtab_2>>=
wlda15 %$% qtab(OECD, income, w = POP) %>% divide_by(1e6)
@
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e. \fct{fmean} is only invoked once).
<<qtab_3>>=
wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean) %>% replace_na(0)
@
Finally, this calculates a population-weighted average of life expectancy in each group.
<<qtab_4>>=
wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean,
                wFUN.args = list(w = POP)) %>% replace_na(0)
@
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above functions, \pkg{collapse} also provides functions \pkg{pwcor}, \pkg{pwcov}, \pkg{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, and also functions \pkg{psacf}, \pkg{pspacf} and \pkg{psccf} for auto- and cross-covariance and correlation function estimation on panel series.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op} (for safety and performance reasons) visible in the \href{https://sebkrantz.github.io/collapse/reference/fmean.html}{documentation of some functions}. Its contents can be accessed using \fct{get\_collapse}. \newline

>>>>

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

What is the purpose of combining all of this in a package? The short answer is to make computations in \proglang{R} as flexible and powerful as possible. The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline

@@

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars} and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy} or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.}. Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function. However, as demonstrated below, the possibility of combining vectorized statistical functions also permits calculating more complex statistics in a vectorized way. \newline

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of Single Instruction Multiple Data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}. The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there, not here for parsimony) demonstrate their basic usage.

@@

fmean(mtcars$mpg)
fmean(EuStockMarkets)
fmean(mtcars)
fmean(mtcars$mpg, w = mtcars$wt)
fmean(mtcars$mpg, g = mtcars$cyl)
fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")

@@

  3 && \code{"-"}   && subtract (i.e., center) \\

@@

  5 && \code{"/"}   && divide (i.e., scale) \\
  6 && \code{"\%"}     && compute percentages (i.e., divide and multiply by 100) \\

@@

  9 && \code{"\%\%"} && modulus (i.e., remainder from division by  \code{STATS}) \\
  10 && \code{"-\%\%"} && subtract modulus (i.e., make data divisible by \code{STATS})

@@

to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can be an \class{indexed\_frame} with any number of individual and/or time identifiers (e.g., an indexed \class{data.table} is fully functional for other operations). The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}. The basic syntax is:

@@

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \code{base::table}. It is enhanced both in a statistical and computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights - vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
%
<<qtab>>=
library("magrittr") # World after 2015 (latest country data)
wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
wlda15 %$% qtab(OECD, income)
@
This shows the total population (latest post-2015 estimates) in millions.
<<qtab_2>>=
wlda15 %$% qtab(OECD, income, w = POP) %>% divide_by(1e6)
@
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only invoked once).
<<qtab_3>>=
wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean) %>% replace_na(0)
@
Finally, this calculates a population-weighted average of life expectancy in each group.
<<qtab_4>>=
wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean,
                wFUN.args = list(w = POP)) %>% replace_na(0)
@
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above functions, \pkg{collapse} also provides functions \pkg{pwcor}, \pkg{pwcov}, \pkg{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, and also functions \pkg{psacf}, \pkg{pspacf} and \pkg{psccf} for auto- and cross-covariance and correlation function estimation on panel series.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op} (for safety and performance reasons) visible in the \href{https://sebkrantz.github.io/collapse/reference/fmean.html}{documentation of some functions}. Its contents can be accessed using \fct{get\_collapse}. \newline
